---
title: "Metagene Analysis with RMetaTools"
author: "Manfred Schmid"
date: "`r Sys.Date()`"
output: 
      rmarkdown::html_vignette:
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Metagene Analysis with RMetaTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Load Package

```{r}
library('RMetaTools')
```


## Loading Regions

Using a bed file this will create a range around an anchor point *TSS*, *TES* or *center*. Note, under the hood this uses rtracklayer to load, so gtfs etc should work, but non-conventional columns in your files may cause this to fail.

```{r}
bedfile <- system.file("extdata", "Chen_PROMPT_TSSs_liftedTohg38.bed", package = "RMetaTools")

anchor = 'TSS'
upstream = 1000
downstream = 1000
window_size = 50
```

```{r}
regions <- meta_regions(bedfile, 'TSS', upstream, downstream, window_size)

regions
```


## Create a metagene matrix

```{r}
bw_plus <- system.file("extdata", "GSM1573841_mNET_8WG16_siLuc_plus_hg38.bw", package = "RMetaTools")
bw_minus <- system.file("extdata", "GSM1573841_mNET_8WG16_siLuc_minus_hg38.bw", package = "RMetaTools")

mat <- get_matrix(bw_plus, bw_minus, regions, upstream, downstream, window_size)

head(mat)
```


this matrix can be directly plotted either as heatmap or colMeans, etc...
```{r}
heatmap(as.matrix(mat), Rowv = NA, Colv = NA, labRow = NA, labCol = NA)
```


```{r}
plot(colSums(mat), type='l')
```



## tidy metagene analysis

I prefer to use tidyverse logic for further analysis even though that me get problematic for big datasets.

Convert the matrix to fully annotated tidy tbl can be done with the built-in function.
```{r}
(tidy_meta <- mat_to_tbl(mat, regions, anchor, upstream, downstream, window_size))
```

There is also a built-in heatmap plotter for the tidy data
```{r}
plot_htmp(tidy_meta)
```


May be useful to represent as log2 values
```{r}
log2_tidy_meta <- mutate(tidy_meta, value=log2(value+1))

plot_htmp(log2_tidy_meta) +
  scale_fill_gradient(low='white', high='navyblue') + #custom color scheme
  coord_fixed(ratio = 4) #y/x aspect ratio of output plot
```




## metagene profiles

So far this is simply doing a one-sample t.test which computes mean and parametric confidence intervals.
```{r}
log2_tidy_meta_avg <- meta_average(log2_tidy_meta)
```

This can be plotted easily
```{r}
plot_profile(log2_tidy_meta_avg)
```



## Recipe: order heatmap rows using sum of values per row

To order heatmap rows in ggplot2 requires a bit of work.
ie to order according to the sum in each row, we first compute and order '*by-hand*'.
```{r}
value_order <- log2_tidy_meta %>%
  group_by(gene) %>%
  summarize(total_value = sum(value)) %>%
  arrange(total_value) %$%
  gene
```

We can apply this ordering to the tbl using *factor()*.
```{r}
log2_tidy_meta %>%
  mutate(gene = factor(gene, levels=value_order)) %>%
  plot_htmp(., do_interpolate = TRUE) +
  scale_fill_gradient(low='white', high='navyblue')
```
  
  
## Recipe: add another matrix side-by-side

First create the new matrix.
```{r}
bw_plus2 <- system.file("extdata", "GSM1573839_mNET_8WG16_siCPSF73_plus_hg38.bw", package = "RMetaTools")
bw_minus2 <- system.file("extdata", "GSM1573839_mNET_8WG16_siCPSF73_minus_hg38.bw", package = "RMetaTools")

mat2 <- get_matrix(bw_plus2, bw_minus2, regions, upstream, downstream, window_size)

tidy_meta2 <- mat_to_tbl(mat2, regions, anchor, upstream, downstream, window_size)
```


Before we combine, better add some identifier column, ie 'condition'.
```{r}
tidy_meta$condition <- 'ctrl'
tidy_meta2$condition <- 'siCPSF73'
```

simply bind_rows then
```{r}
both_meta <- bind_rows(tidy_meta, tidy_meta2)
log2_both_meta <- mutate(both_meta, value=log2(value+1))
```


#### heatmap side-by-side
```{r}
plot_htmp(log2_both_meta) +
  scale_fill_gradient(low='white', high='navyblue') +
  facet_grid(.~condition)
```


#### profiles of combine dataset

average of a combined dataset, should work
```{r}
log2_both_meta_avg <- meta_average(log2_both_meta)
```

can be visualized as side-by-side plot:
```{r}
plot_profile(log2_both_meta_avg) +
  facet_grid(.~condition)
```


but probably often more informative using overlay:
```{r}
plot_profile(log2_both_meta_avg, color_by='condition')
```



#### Recipe: log2 ratio between conditions

using the tidy tbl this is pretty straightforward:
```{r}
(LFCs <- log2_both_meta %>%
  spread(condition, value) %>%
  mutate(value = siCPSF73-ctrl) %>%
  dplyr::select(-siCPSF73, -ctrl))
```

```{r}
plot_htmp(LFCs, color_by='value') +
  scale_fill_gradient2(low='firebrick4', mid='white', high='navyblue', name="log2(siCPSF73/ctrl)") +
  coord_fixed(ratio = 4) 
```


```{r}
ratio_profile <- meta_average(LFCs)
```

```{r}
plot_profile(ratio_profile) +
  ylab('mean log2FC siCPSF73/ctrl')
```
