---
title: "Metagene Analysis with RMetaTools"
author: "Manfred Schmid"
date: "`r Sys.Date()`"
output: 
      rmarkdown::html_vignette:
        fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Metagene Analysis with RMetaTools}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```



# Install and Load Package

usually this should work:
```{r, eval=FALSE}
library(devtools)
install_github('manschmi/RMetaTools') #full install including dependencies
```

and then this should not throw an error
```{r}
library(RMetaTools)
```


#### INSTALL on ie genome.dk

Handling R packages at genome.dk can be tricky, ie installing devtools which will be needed failed for me so better stick to 'conda' for handling such annoyancies. Ups: you will have to install conda as described in the genome.au.dk helppages.

Run this in the Terminal, front-end: (ups can take a long time each step...). Just say *yes* to everything when asked.
```{bash, eval=F}
conda create -n RMetaTools_env
conda activate RMetaTools_env
conda install -c r r
conda install -c r rstudio
conda install -c r r-devtools
conda install -c bioconda bioconductor-rtracklayer
```

now go to backend activate the environment again and start R
```{bash, eval=FALSE}
srun --pty bash
conda activate RMetaTools_env
rstudio #or R if you prefer that, rstudio requires that you logged in with ssh -Y
```


install RMetaTools for use in RMetaTools_env at genome.dk:
```{r, eval=FALSE}
library(devtools)
library(httr)
set_config( config( ssl_verifypeer = 0L ) )
options(unzip = "internal")

install_github('manschmi/RMetaTools') #full install including dependencies
install_github('manschmi/RMetaTools', upgrade=FALSE) #only RMetaTools but no dependencies are updated
```


check by trying to load RMetaTools
```{r}
library('RMetaTools')
```

Hopefully this does not throw and error.
Now you should be all set to try and runt he examples below and/or test with your own data!


# Loading and visualizing deeptools matrices


This should work with any deeptools matrix. From deeptools v2 to v3 some things changed so there are 2 versions.


```{r}
?load_deeptoolsmatrix3
```

#### for deeptools v2

probably never used anymore
```{r}
fname <- system.file("extdata", "matrix.gz", package = "RMetaTools")
df <- load_deeptoolsmatrix(fname)

df
```

#### for deeptools v3
```{r}
fname <- system.file("extdata", "deeptools3_matrix.gz", package = "RMetaTools")

df <- load_deeptoolsmatrix3(fname)
```


#### simple metagene plot therefrom

ups: this example has way too many samples --> simplify before plotting
```{r}
mini_df <- df %>%
  filter(grepl('^EGFP', sample_name) | grepl('^RRP40', sample_name)) %>%
  mutate(sample_name = sub('_hg38.*', '', sample_name))
```

```{r}
mini_df %>%
  group_by(sample_name, rel_pos) %>%
  summarize(mean_log2_value = mean(log2(value+1))) %>%
  ggplot(., aes(x=rel_pos, y=mean_log2_value, color=sample_name)) +
  geom_line() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```



##more funky metagene profile with confidence interval
needs a little extra package for tidy t-testing (should be loaded but just to make sure)

```{r}
library('broom')
```


```{r}
(meta_values <- mini_df %>%
  group_by(sample_name, rel_pos) %>%
  do(tidy(t.test(log2(.$value+1)))))
```

```{r}
meta_values %>%
  ggplot(.) +
  geom_line(aes(x=rel_pos, y=estimate, color=sample_name)) +
  geom_ribbon(aes(x=rel_pos, ymin=conf.low, ymax=conf.high, fill=sample_name, stroke=NA), alpha=.2) +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```


#### a heatmap

```{r}
mini_df %>%
  ggplot(., aes(x=rel_pos, y=id, fill=log2(value+1))) +
  geom_tile() +
  facet_grid(.~sample_name) +
  theme(axis.text.y = element_blank()) +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

#### a heatmap sorted by signal

```{r}
ids_sum_sorted <- mini_df %>%
  group_by(id) %>%
  summarize(total_value = sum(value)) %>%
  arrange(total_value) %$%
  id
```

```{r}
mini_df %>%
  mutate(id = factor(id, levels=ids_sum_sorted)) %>%
  ggplot(., aes(x=rel_pos, y=id, fill=log2(value+1))) +
  geom_tile() +
  facet_grid(.~sample_name) +
  theme(axis.text.y = element_blank()) +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```



# Metagene Analysis Directly with RMetaTools


#### Load metagene values

First step is to define the regions of interest.

Using a bed file this tool will query a range around an anchor point *TSS*, *TES* or *center*. Note, under the hood this uses rtracklayer to load, so gtfs etc should work, but non-conventional columns in your files may cause this to fail.

```{r}
bedfile <- system.file("extdata", "Chen_PROMPT_TSSs_liftedTohg38.bed", package = "RMetaTools")

anchor = 'TSS'
upstream = 1000
downstream = 1000
window_size = 50
```


Track files, this are usually 2 bigwig files for stranded data. If you are assaying unstranded data use the same bigwig file for bw_plus and bw_minus.

```{r}
bw_plus <- system.file("extdata", "GSM1573841_mNET_8WG16_siLuc_plus_hg38.bw", package = "RMetaTools")
bw_minus <- system.file("extdata", "GSM1573841_mNET_8WG16_siLuc_minus_hg38.bw", package = "RMetaTools")

meta_tbl <- RMetaTools::metagene_matrix(bw_plus, bw_minus, bedfile, anchor, upstream, downstream, window_size)

meta_tbl
```




#### Plot metagene values

There is also a built-in heatmap plotter for the tidy data
```{r}
plot_htmp(meta_tbl)
```


May be useful to represent as log2 values
```{r}
(pseudocount <- min(meta_tbl$value[meta_tbl$value > 0]))
```


```{r}
log2_meta_tbl <- mutate(meta_tbl, value=log2(value+pseudocount))
```

```{r}
plot_htmp(log2_meta_tbl) +
  scale_fill_gradient(low='white', high='navyblue') + #custom color scheme
  coord_fixed(ratio = 4) +#y/x aspect ratio of output plot
  theme(axis.text.x = element_text(angle=45, hjust=1), #rotate x axis labels
        axis.title.y = element_blank()) +#remove y axis title
  xlab('bp to PROMPT TSS')
```

this is fine, the only caveat here is that negatives are mentally a bit challenging, ie they still refer to "positive signal".

so its much more paedogical to use pseudocount = 1

```{r}
log2_meta_tbl <- mutate(meta_tbl, value=log2(value+1))
```

```{r}
plot_htmp(log2_meta_tbl) +
  scale_fill_gradient(low='white', high='navyblue') + #custom color scheme
  coord_fixed(ratio = 4) +#y/x aspect ratio of output plot
  theme(axis.text.x = element_text(angle=45, hjust=1), #rotate x axis labels
        axis.title.y = element_blank()) +#remove y axis title
  xlab('bp to PROMPT TSS')
```

but obviously this also changes the appearance quite a bit, since low values are now very close to 0... matter of choice imho.



## metagene profiles

So far this is simply doing a one-sample t.test which computes mean and parametric confidence intervals.
```{r}
log2_meta_tbl_avgs <- meta_average(log2_meta_tbl)
```

This can be plotted easily
```{r}
plot_profile(log2_meta_tbl_avgs)
```




# Recipes

## using Rscript metageneR.R

Instead of running the metagene matrix generation interactively, it can alos be run using *Rscript*. The easiest is to get this from github at \link{https://raw.githubusercontent.com/manschmi/RMetaTools/master/scripts/metageneR.R}.

or on genome.au.dk in the Terminal:
```{bash, eval=FALSE}
wget https://raw.githubusercontent.com/manschmi/RMetaTools/master/scripts/metageneR.R
```

which will add metageneR.R to the current location.

in a Terminal you can now do:
```{bash, eval=FALSE}
Rscript metageneR.R -h
```



## Recipe: order heatmap rows 
#### using sum of values per row

To order heatmap rows in ggplot2 requires a bit of work.
ie to order according to the sum in each row, we first compute and order '*by-hand*'.
```{r}
value_order <- log2_meta_tbl %>%
  filter(relpos == 0 | relpos==50) %>% #optional: select ie first 2 50bp intervals from TSS
  group_by(name) %>%
  summarize(total_value = sum(value)) %>%
  arrange(total_value) %$%
  name
```

We can apply this ordering to the tbl using *factor()*.
```{r}
log2_meta_tbl %>%
  mutate(name = factor(name, levels=value_order)) %>%
  plot_htmp(., do_interpolate = TRUE) +
  scale_fill_gradient(low='white', high='navyblue')
```
  

#### using region size

ups for PROMPT annotation used above we only have TSS position but no actual size from start. For this better to use something else, ie snRNA.
```{r}
snRNA_bedfile <- system.file("extdata", "snRNA_MS31_hg38.bed", package = "RMetaTools")
```

```{r}
snRNA_meta_tbl <- RMetaTools::metagene_matrix(bw_plus, bw_minus, snRNA_bedfile, anchor, upstream, downstream, window_size) %>%
  mutate(value = log2(value+1))

snRNA_meta_tbl
```

without ordering
```{r}
plot_htmp(snRNA_meta_tbl, do_interpolate = TRUE)
```

The actual annotations are never used in the default pipeline but you can get the info from the name.
```{r}
size_order <- snRNA_meta_tbl %>%
  distinct(name) %>%
  separate(name, c('chr', 'start', 'end'), sep='\t', extra='drop', remove = FALSE) %>%
  mutate(region_size = as.integer(end) - as.integer(start)) %>%
  dplyr::select(name, region_size) %>%
  arrange(-region_size)
```

Then apply the order as factor levels as above.
```{r}
p <- snRNA_meta_tbl %>%
  mutate(name = factor(name, levels=size_order$name)) %>%
  plot_htmp(., do_interpolate = TRUE)

p
```

and add lines for anchor and regions size (ups: not tested rigorously):
```{r}
p +
  geom_vline(xintercept = 0, size=.1, color='black', alpha=.5) +
  geom_point(data = mutate(size_order, value=0), aes(x=region_size, y=name), size=1, color='black', alpha=.5, stroke=0)
```

## Recipe: add another matrix side-by-side

First create the new matrix.
```{r}
bw_plus2 <- system.file("extdata", "GSM1573839_mNET_8WG16_siCPSF73_plus_hg38.bw", package = "RMetaTools")
bw_minus2 <- system.file("extdata", "GSM1573839_mNET_8WG16_siCPSF73_minus_hg38.bw", package = "RMetaTools")

meta_tbl2 <- metagene_matrix(bw_plus2, bw_minus2, bedfile, anchor, upstream, downstream, window_size)

```


Before we combine, better add some identifier column, ie 'condition'.
```{r}
meta_tbl$condition <- 'ctrl'
meta_tbl2$condition <- 'siCPSF73'
```

simply bind_rows then
```{r}
both_meta <- bind_rows(meta_tbl, meta_tbl2)
log2_both_meta <- mutate(both_meta, value=log2(value+1))
```


#### heatmap side-by-side
```{r}
log2_both_meta %>%
  mutate(name = factor(name, levels=value_order)) %>%
plot_htmp(.) +
  scale_fill_gradient(low='white', high='navyblue') +
  facet_grid(.~condition)
```


#### profiles of combine dataset

average of a combined dataset, should work
```{r}
log2_both_meta_avg <- meta_average(log2_both_meta)
```

can be visualized as side-by-side plot:
```{r}
plot_profile(log2_both_meta_avg) +
  facet_grid(.~condition)
```


but probably often more informative using overlay:
```{r}
plot_profile(log2_both_meta_avg, color_by='condition')
```





## Recipe: Add antisense coverage for comparison "on the bottom"

For this example we simply invert plus_bw and minus_bw --> ie get the antisense coverage on same annotations. Alternatively you can use other bedfile but same bws as above...

First create the new matrix to be put on the bottom. We use the set with already 2 columns ;-)

```{r}
bottom1 <- metagene_matrix(bw_minus, bw_plus, bedfile, anchor, upstream, downstream, window_size)  %>%
  mutate(condition='ctrl')

bottom2 <- metagene_matrix(bw_minus2, bw_plus2, bedfile, anchor, upstream, downstream, window_size) %>%
  mutate(condition='siCPSF73')

bottom <- bind_rows(bottom1, bottom2) %>%
  mutate(strand = 'antisense')
```

```{r}
log2_meta_all <- both_meta %>%
  mutate(strand = 'sense') %>%
  bind_rows(., bottom) %>%
  mutate(value=log2(value+1),
         strand=factor(strand, levels=c('sense', 'antisense')))
```


#### heatmap 2x2 matrix
```{r}
log2_meta_all %>%
  mutate(name = factor(name, levels=value_order),
         value = ifelse(strand == 'sense', value, -value)) %>%
plot_htmp(.) +
  scale_fill_gradient2(low='firebrick', mid='white', high='navyblue') +
  facet_grid(strand~condition) +
  theme_minimal() +
  theme(axis.text.y=element_blank(),
        axis.title.y = element_blank())
```

#### heatmap sense antisense color overlay

this is a bit iffy but can be faked by duplicate each rowname and offsetting accordingly

rows of the heatmap are ordered by levels of 'name'. So we simply duplicate each name and order so that the positive strand values for each name are always followed by the negative strand values for the same name .

This is the order we apply now:
```{r}
head(value_order)
```

lets make a value order for antisense


put this 2 together zipped way

```{r}
zipped_value_order <- as.character(rbind(value_order,paste0(value_order, 'antisense')))

head(zipped_value_order)
```


```{r}
log2_meta_all %>%
  mutate(name = ifelse(strand == 'antisense', name, paste0(value_order, 'antisense')),
         name = factor(name, levels=zipped_value_order),
         value = ifelse(strand == 'sense', value, -value)) %>%
plot_htmp(.) +
  scale_fill_gradient2(low='firebrick', mid='white', high='navyblue') +
  facet_grid(.~condition) +
  theme_minimal() +
  theme(axis.text.y=element_blank(),
        axis.title.y = element_blank())
```

another version would simply use a majority vote for each tile

```{r}
log2_meta_all %>%
  spread(strand, value) %>%
  mutate(value = ifelse(sense > antisense, sense, -antisense),
         name = factor(name, levels=zipped_value_order)) %>%
plot_htmp(.) +
  scale_fill_gradient2(low='firebrick', mid='white', high='navyblue') +
  facet_grid(.~condition) +
  theme_minimal() +
  theme(axis.text.y=element_blank(),
        axis.title.y = element_blank())
```



#### profiles of 2x2 dataset

average of a combined dataset, should work
```{r}
log2_meta_all_avg <- meta_average(log2_meta_all)
```

can be visualized as side-by-side plot:
```{r}
plot_profile(log2_meta_all_avg) +
  facet_grid(strand~condition)
```


#### 2x2 Strand vs antisense compound plot:

```{r}
log2_meta_all_avg %>%
  mutate(estimate = ifelse(strand == 'antisense', -estimate, estimate),
         conf.low = ifelse(strand == 'antisense', -conf.low, conf.low),
         conf.high = ifelse(strand == 'antisense', -conf.high, conf.high)) %>%
plot_profile(., color_by='condition') +
  facet_grid(strand~., scales='free') +
  theme_minimal()
```




## Recipe: log2 ratio between conditions

using the tidy tbl this is pretty straightforward:
```{r}
(LFCs <- log2_both_meta %>%
  spread(condition, value) %>%
  mutate(value = siCPSF73-ctrl) %>%
  dplyr::select(-siCPSF73, -ctrl))
```

```{r}
plot_htmp(LFCs, color_by='value') +
  scale_fill_gradient2(low='firebrick4', mid='white', high='navyblue', name="log2(siCPSF73/ctrl)") +
  coord_fixed(ratio = 4) 
```


```{r}
ratio_profile <- meta_average(LFCs)
```

```{r}
plot_profile(ratio_profile) +
  ylab('mean log2FC siCPSF73/ctrl')
```



## Recipe: sensitivity (Sandelin lab style) ratio between conditions instead of log2FC

using the tidy tbl this is pretty straightforward:
```{r}
(pseudocount <- min(both_meta$value[both_meta$value > 0]))

```

```{r}
(sensitivities <- both_meta %>%
  mutate(value = value + pseudocount) %>%
  spread(condition, value) %>%
  mutate(value = (siCPSF73-ctrl)/max(siCPSF73, ctrl)) %>%
  dplyr::select(-siCPSF73, -ctrl))
```

```{r}
plot_htmp(sensitivities, color_by='value') +
  scale_fill_gradient2(low='firebrick4', mid='white', high='navyblue', name="sensitivity = (siCPSF73-ctrl)/max(siCPSF73, ctrl)") +
  coord_fixed(ratio = 4) 
```


```{r}
sensitivity_profile <- meta_average(sensitivities)
```

```{r}
plot_profile(sensitivity_profile) +
  ylab('mean sensitibity siCPSF73/ctrl')
```


## Recipe: p value of log2 ratio between conditions

using a t.test would require roughly normal distributed data. This is often not the case due to highly 0-inflated data. As shown here for the density of signals in the central bins. The TSS and 50bp bin are close to normal, but the others are highly skewed towards 0s.

```{r}
log2_both_meta %>%
  filter(relpos > -100 & relpos < 200) %>%
  mutate(relpos = factor(relpos)) %>%
  ggplot(., aes(x=value, color=relpos)) +
  geom_density() +
  facet_grid(.~condition) +
  theme_minimal()
```

So its probably more appropriate to do a non-parametric test ie wilcoxon rank sum test. Together with the tidy function from R package *broom*.

```{r}
suppressWarnings(library(broom))
```


```{r}
(LFC_wilcox_p <- log2_both_meta %>%
  spread(condition, value) %>%
  group_by(relpos) %>%
  do(tidy(wilcox.test(.$siCPSF73, .$ctrl))))
```

This applies a significance test for every position. Hence, p-values need to be corrected for multiple testing:

```{r}
LFC_wilcox_p %<>%
  mutate(padj = p.adjust(p.value))
```


plot the adjusted p values, ie visually better as -log10(p.value)

```{r}
ggplot(LFC_wilcox_p,
       aes(x=relpos, y=-log10(padj))) +
  geom_line()
```


## Recipe: bigwigs where minus bigwig values are represented as negatives

Some people present values of minus strand data as negatives. This can help visualization in Genome Browser, etc.. To make such datasets directly accessible I added an option to negate negative strand bigwigs: 
```{r, eval=FALSE}
meta_tbl <- RMetaTools::metagene_matrix(bw_plus, bw_minus, bedfile, anchor, upstream, downstream, window_size, negate_neg_strand_values = TRUE)
```


# Individual Steps in the Pipeline

## Loading Regions

Using a bed file this will create a range around an anchor point *TSS*, *TES* or *center*. Note, under the hood this uses rtracklayer to load, so gtfs etc should work, but non-conventional columns in your files may cause this to fail.

```{r}
bedfile <- system.file("extdata", "Chen_PROMPT_TSSs_liftedTohg38.bed", package = "RMetaTools")

anchor = 'TSS'
upstream = 1000
downstream = 1000
window_size = 50
```

```{r}
regions <- meta_regions(bedfile, 'TSS', upstream, downstream)

regions
```


## Create a metagene matrix

On top of the general method one can also get the intermediate non-tidy matrix separately:
```{r}
bw_plus <- system.file("extdata", "GSM1573841_mNET_8WG16_siLuc_plus_hg38.bw", package = "RMetaTools")
bw_minus <- system.file("extdata", "GSM1573841_mNET_8WG16_siLuc_minus_hg38.bw", package = "RMetaTools")

mat <- get_matrix(bw_plus, bw_minus, regions, upstream, downstream, window_size)

head(mat)
```


this matrix can be directly plotted either as heatmap or colMeans, etc...
```{r}
heatmap(as.matrix(mat), Rowv = NA, Colv = NA, labRow = NA, labCol = NA)
```


```{r}
plot(colSums(mat), type='l')
```


Conversion to tidy format is also implemented as function. This should return exactly same data as the metagene_matrix functioncreated using the same settings.
```{r}
(tidy_meta <- mat_to_tbl(mat))
```



# Session

```{r}
sessionInfo()
```

